\chapter{Operator Semigroups and Hille--Yosida}

\section{Strongly continuous semigroups and generators}

\begin{notedefinition}
Let $X$ be a Banach space.  A family $(T(t))_{t\ge0}\subseteq\Bop{X}{X}$ is a
\emph{strongly continuous semigroup}, or $C_0$-semigroup, if
\[
  T(0)=I,
  \qquad
  T(t+s)=T(t)T(s),
\]
and
\[
  \lim_{t\downarrow0}T(t)x=x
  \qquad(x\in X).
\]
Its \emph{infinitesimal generator} is
\[
  Ax=\lim_{t\downarrow0}\frac{T(t)x-x}{t},
\]
with domain
\[
  \Dom(A)=\left\{x\in X:
       \lim_{t\downarrow0}\frac{T(t)x-x}{t}
       \text{ exists in }X\right\}.
\]
\end{notedefinition}

\begin{noteproposition}
Let $A$ be the generator of a $C_0$-semigroup $(T(t))$.  Then:
\begin{enumerate}
  \item $A$ is linear and $\Dom(A)$ is dense in $X$;
  \item $T(t)\Dom(A)\subseteq\Dom(A)$ and
        $AT(t)x=T(t)Ax$ for $x\in\Dom(A)$;
  \item for $x\in\Dom(A)$, the orbit is continuously differentiable and
        \[
          \frac{\dd}{\dd t}T(t)x=AT(t)x=T(t)Ax;
        \]
  \item $A$ is closed.
\end{enumerate}
\end{noteproposition}

\begin{proof}
Linearity follows from the limit.  For density, set
\[
  x_h=\frac1h\int_0^hT(s)x\,\dd s.
\]
Strong continuity gives $x_h\to x$, while
\[
  \frac{T(t)x_h-x_h}{t}
  =\frac1h\left(
       \frac1t\int_h^{h+t}T(s)x\,\dd s
       -\frac1t\int_0^tT(s)x\,\dd s
    \right)
  \longrightarrow\frac{T(h)x-x}{h}.
\]
Thus $x_h\in\Dom(A)$.

For $x\in\Dom(A)$,
\[
 \frac{T(h)T(t)x-T(t)x}{h}
 =T(t)\frac{T(h)x-x}{h}\longrightarrow T(t)Ax,
\]
which proves invariance and commutation.  The derivative at arbitrary $t$
follows from the semigroup law; continuity of $s\mapsto T(s)Ax$ makes it a
continuous derivative.

Finally, suppose $x_n\to x$ and $Ax_n\to y$.  The fundamental theorem of
calculus gives
\[
  T(t)x_n-x_n=\int_0^tT(s)Ax_n\,\dd s.
\]
Passing to the limit yields
\[
  T(t)x-x=\int_0^tT(s)y\,\dd s.
\]
Divide by $t$ and let $t\downarrow0$ to obtain $x\in\Dom(A)$ and $Ax=y$.
\end{proof}

\begin{notedefinition}
A $C_0$-semigroup is a \emph{contraction semigroup} if
$\|T(t)\|\le1$ for every $t\ge0$.
\end{notedefinition}

\section{Resolvents of generators}

\begin{notelemma}[Laplace-transform resolvent]
Let $(T(t))$ be a contraction semigroup with generator $A$.  For
$\operatorname{Re}\lambda>0$, define
\[
  R(\lambda,A)x
  =\int_0^\infty e^{-\lambda t}T(t)x\,\dd t.
\]
Then the Bochner integral is well defined,
\[
  R(\lambda,A)=(\lambda I-A)^{-1},
  \qquad
  \|R(\lambda,A)\|\le\frac1{\operatorname{Re}\lambda}.
\]
For real $\lambda>0$,
\[
  \|\lambda R(\lambda,A)\|\le1.
\]
\end{notelemma}

\begin{proof}
The norm estimate follows from
\[
 \int_0^\infty e^{-\operatorname{Re}\lambda t}\|T(t)x\|\,\dd t
 \le\frac{\|x\|}{\operatorname{Re}\lambda}.
\]
For $s>0$, the semigroup law and a change of variables give
\[
 \frac{T(s)-I}{s}R(\lambda,A)x
 =\frac1s\int_0^\infty e^{-\lambda t}
       \bigl(T(t+s)-T(t)\bigr)x\,\dd t
 \longrightarrow \lambda R(\lambda,A)x-x.
\]
Hence $R(\lambda,A)x\in\Dom(A)$ and
$(\lambda I-A)R(\lambda,A)x=x$.  If $x\in\Dom(A)$, integration by parts in
$e^{-\lambda t}T(t)x$ gives
$R(\lambda,A)(\lambda I-A)x=x$.  Thus it is the inverse.
\end{proof}

\begin{notecorollary}
The generator of a contraction semigroup is densely defined and closed,
$(0,\infty)\subseteq\rho(A)$, and
\[
  \|\lambda(\lambda I-A)^{-1}\|\le1
  \qquad(\lambda>0).
\]
\end{notecorollary}

\section{The contraction Hille--Yosida theorem}

\begin{notetheorem}[Hille--Yosida theorem, contraction form]
Let $A\colon\Dom(A)\subseteq X\to X$ be linear.  Then $A$ generates a
contraction $C_0$-semigroup if and only if
\begin{enumerate}
  \item $A$ is densely defined and closed;
  \item $(0,\infty)\subseteq\rho(A)$;
  \item $\|\lambda(\lambda I-A)^{-1}\|\le1$ for every $\lambda>0$.
\end{enumerate}
\end{notetheorem}

\begin{proof}
Necessity is exactly the Laplace-transform resolvent lemma, together with the
general facts that a generator is densely defined and closed.  We prove
sufficiency.

Write $R_\lambda=(\lambda I-A)^{-1}$ and, for $\lambda>0$, define the bounded
Yosida approximation
\[
  A_\lambda
  =\lambda AR_\lambda
  =\lambda^2R_\lambda-\lambda I.
\]
The bounded operator $A_\lambda$ generates the uniformly continuous semigroup
\[
  T_\lambda(t)=e^{tA_\lambda}.
\]
Since $\|\lambda R_\lambda\|\le1$,
\[
\begin{aligned}
  \|T_\lambda(t)\|
   &=\left\|e^{-\lambda t}
      \sum_{n=0}^\infty\frac{(\lambda^2t)^n}{n!}R_\lambda^n\right\|\\
   &\le e^{-\lambda t}
      \sum_{n=0}^\infty\frac{(\lambda t)^n}{n!}=1.
\end{aligned}
\]
Thus all approximating semigroups are contractions.

We next record the approximation property.  If $y\in\Dom(A)$, then
\[
  \lambda R_\lambda y-y=R_\lambda Ay,
\]
so $\lambda R_\lambda y\to y$.  Since $\|\lambda R_\lambda\|\le1$ and
$\Dom(A)$ is dense, the same convergence holds for every $y\in X$.  Therefore,
for $x\in\Dom(A)$,
\[
  A_\lambda x=\lambda R_\lambda Ax\longrightarrow Ax.
\]
All resolvents commute, and hence the operators $A_\lambda$ commute.  For
$x\in\Dom(A)$, the bounded-operator variation formula gives
\[
\begin{aligned}
 T_\lambda(t)x-T_\mu(t)x
   &=\int_0^t
      T_\lambda(t-s)T_\mu(s)(A_\lambda-A_\mu)x\,\dd s,
\end{aligned}
\]
whence
\[
  \sup_{0\le t\le t_0}
  \|T_\lambda(t)x-T_\mu(t)x\|
  \le t_0\|(A_\lambda-A_\mu)x\|\longrightarrow0.
\]
For arbitrary $x\in X$, approximate $x$ by an element of $\Dom(A)$ and use
the contraction bounds.  Thus, for each $x$, $(T_\lambda(t)x)$ is Cauchy
uniformly on compact time intervals.  Define
\[
  T(t)x=\lim_{\lambda\to\infty}T_\lambda(t)x.
\]
Passing to the limit in
$T_\lambda(t+s)=T_\lambda(t)T_\lambda(s)$ proves the semigroup law, and the
norm estimate passes to the limit.  Strong continuity follows first on
$\Dom(A)$ from
\[
  T_\lambda(t)x-x=\int_0^tT_\lambda(s)A_\lambda x\,\dd s
\]
and the uniform boundedness of $A_\lambda x$ for large $\lambda$, and then on
all of $X$ by density and the contraction bound.

Let $G$ be the generator of $(T(t))$.  For $x\in\Dom(A)$, uniform convergence
on compact time intervals, $A_\lambda x\to Ax$, and the contraction estimate
allow passage to the limit in the last integral:
\[
  T(t)x-x=\int_0^tT(s)Ax\,\dd s.
\]
Dividing by $t$ and letting $t\downarrow0$ gives
$x\in\Dom(G)$ and $Gx=Ax$; hence $A\subseteq G$.

To prove equality, fix $\mu>0$.  Both $\mu I-A$ and $\mu I-G$ are bijective.
For $y\in X$, put $x=R(\mu,A)y$.  Then $x\in\Dom(A)\subseteq\Dom(G)$ and
\[
  (\mu I-G)x=(\mu I-A)x=y.
\]
Thus $R(\mu,G)y=R(\mu,A)y$ for every $y$.  Equal resolvents have the same
range, so $\Dom(G)=\Dom(A)$, and then $G=A$.
\end{proof}

\begin{notecorollary}[Uniqueness and the abstract Cauchy problem]
A generator determines its $C_0$-semigroup uniquely.  If $x_0\in\Dom(A)$,
then
\[
  u(t)=T(t)x_0
\]
is the unique classical solution of
\[
  u'(t)=Au(t),
  \qquad u(0)=x_0.
\]
Moreover, $\Dom(A^2)$ is a core for $A$ and hence is dense in $\Dom(A)$ for
the graph norm.
\end{notecorollary}

\begin{proof}
Existence follows from the orbit differentiation formula.  If $u$ is another
classical solution, fix $t>0$ and define $F(s)=T(t-s)u(s)$ for $0\le s\le t$.
Since $u(s)\in\Dom(A)$ and $u'(s)=Au(s)$,
\[
  F'(s)=-AT(t-s)u(s)+T(t-s)Au(s)=0.
\]
Hence $u(t)=F(t)=F(0)=T(t)u(0)$, proving uniqueness and, consequently,
uniqueness of the semigroup generated by $A$.

For the core assertion, let $x\in\Dom(A)$ and put
\[
  x_\lambda=\lambda R(\lambda,A)x.
\]
The resolvent maps $X$ into $\Dom(A)$ and commutes with $A$ on $\Dom(A)$, so
\[
  x_\lambda\in\Dom(A^2),
  \qquad
  Ax_\lambda=\lambda R(\lambda,A)Ax.
\]
The resolvent approximation gives
\[
  x_\lambda\to x,
  \qquad
  Ax_\lambda\to Ax.
\]
Thus $x_\lambda\to x$ in the graph norm, proving that $\Dom(A^2)$ is a core.
\end{proof}

\begin{notetheorem}[Euler approximation]
If $A$ generates a contraction semigroup, then for every $x\in X$ and $t\ge0$,
\[
  T(t)x
  =\lim_{n\to\infty}
     \left(I-\frac{t}{n}A\right)^{-n}x.
\]
\end{notetheorem}

\begin{proof}
The case $t=0$ is immediate.  Fix $t>0$ and put $\lambda_n=n/t$.  Repeated use
of the Laplace resolvent formula and the semigroup law gives
\[
  R(\lambda_n,A)^n x
  =\frac1{(n-1)!}\int_0^\infty
       s^{n-1}e^{-\lambda_ns}T(s)x\,\dd s.
\]
Therefore
\[
\begin{aligned}
  \left(I-\frac tnA\right)^{-n}x
   &=(\lambda_nR(\lambda_n,A))^n x\\
   &=\int_0^\infty k_{n,t}(s)T(s)x\,\dd s,
\end{aligned}
\]
where
\[
  k_{n,t}(s)
  =\frac{\lambda_n^n}{(n-1)!}s^{n-1}e^{-\lambda_ns}
\]
is the density of a Gamma random variable $S_n$ with
\[
  \mathbb E S_n=t,
  \qquad
  \operatorname{Var}(S_n)=\frac{t^2}{n}.
\]
The density has integral one.  Hence
\[
  \left\|\left(I-\frac tnA\right)^{-n}x-T(t)x\right\|
  \le\int_0^\infty k_{n,t}(s)\|T(s)x-T(t)x\|\,\dd s.
\]
Given $\varepsilon>0$, strong continuity gives $\delta>0$ such that the
integrand is $<\varepsilon$ whenever $|s-t|<\delta$.  On the complement it is
at most $2\|x\|$, because $T$ is contractive.  Chebyshev's inequality gives
\[
  \int_{|s-t|\ge\delta}k_{n,t}(s)\,\dd s
  \le\frac{t^2}{n\delta^2}\longrightarrow0.
\]
Thus the displayed norm tends to zero, proving the Euler formula.
\end{proof}

\section{Exponentially bounded semigroups}

\begin{notelemma}
Every $C_0$-semigroup is exponentially bounded: there are $M\ge1$ and
$\omega\in\mathbb R$ such that
\[
  \|T(t)\|\le Me^{\omega t}
  \qquad(t\ge0).
\]
\end{notelemma}

\begin{proof}
Strong continuity and uniform boundedness give
$M_0=\sup_{0\le t\le1}\|T(t)\|<\infty$.  Writing $t=n+s$ with
$n\in\mathbb N_0$ and $0\le s<1$ gives, by the semigroup law,
\[
  \|T(t)\|
  =\|T(1)^nT(s)\|
  \le M_0^{n+1}.
\]
Because $T(0)=I$, one has $M_0\ge1$.  Taking
\[
  M=M_0,
  \qquad
  \omega=\log M_0
\]
and using $n\le t$ yields
\[
  \|T(t)\|\le M_0^{n+1}
  \le M_0e^{t\log M_0}=Me^{\omega t}.
\]
\end{proof}

\begin{notelemma}[Higher resolvent powers]
If $\|T(t)\|\le Me^{\omega t}$ and $A$ is its generator, then for
$\operatorname{Re}\lambda>\omega$ and $n\ge1$,
\[
  R(\lambda,A)^n x
  =\frac1{(n-1)!}
    \int_0^\infty t^{n-1}e^{-\lambda t}T(t)x\,\dd t,
\]
and therefore
\[
  \|R(\lambda,A)^n\|
  \le\frac{M}{(\operatorname{Re}\lambda-\omega)^n}.
\]
\end{notelemma}

\begin{proof}
The formula for $n=1$ is the Laplace representation.  Suppose it holds for
$n$.  Since
\[
  \int_0^\infty e^{-\operatorname{Re}\lambda t}\|T(t)x\|\,\dd t
  \le M\|x\|\int_0^\infty
       e^{-(\operatorname{Re}\lambda-\omega)t}\,\dd t<\infty,
\]
the relevant Bochner integrals are absolutely integrable and Fubini's theorem
applies.  Using $T(s)T(t)=T(s+t)$,
\[
\begin{aligned}
 R(\lambda,A)^{n+1}x
 &=\frac1{(n-1)!}
   \int_0^\infty\int_0^\infty
      t^{n-1}e^{-\lambda(s+t)}T(s+t)x\,\dd s\,\dd t\\
 &=\frac1{(n-1)!}
   \int_0^\infty e^{-\lambda r}T(r)x
      \left(\int_0^r t^{n-1}\,\dd t\right)\dd r\\
 &=\frac1{n!}\int_0^\infty
      r^ne^{-\lambda r}T(r)x\,\dd r.
\end{aligned}
\]
This proves the representation by induction.  Taking norms and putting
$a=\operatorname{Re}\lambda-\omega>0$ gives
\[
\begin{aligned}
  \|R(\lambda,A)^nx\|
  &\le\frac{M\|x\|}{(n-1)!}
       \int_0^\infty t^{n-1}e^{-at}\,\dd t\\
  &=\frac{M}{a^n}\|x\|,
\end{aligned}
\]
because the last integral equals $(n-1)!/a^n$.
\end{proof}

\begin{notetheorem}[Hille--Yosida--Phillips theorem]
Let $A$ be densely defined and closed.  It generates a $C_0$-semigroup
satisfying
\[
  \|T(t)\|\le Me^{\omega t}
\]
if and only if $(\omega,\infty)\subseteq\rho(A)$ and
\[
  \|R(\lambda,A)^n\|
  \le\frac{M}{(\lambda-\omega)^n}
  \qquad(\lambda>\omega,\ n\ge1).
\]
\end{notetheorem}

\begin{proof}
Necessity is the higher-resolvent-power lemma.  We prove sufficiency by the
Yosida construction used in the handwritten proof.

For $\lambda>\max\{\omega,0\}$, write
$R_\lambda=(\lambda I-A)^{-1}$ and define
\[
  B_\lambda
  =\lambda AR_\lambda
  =\lambda^2R_\lambda-\lambda I,
  \qquad
  T_\lambda(t)=e^{tB_\lambda}.
\]
The assumed power estimates give
\[
\begin{aligned}
  \|T_\lambda(t)\|
   &=\left\|e^{-\lambda t}
       \sum_{n=0}^\infty\frac{(\lambda^2t)^n}{n!}R_\lambda^n\right\|\\
   &\le M e^{-\lambda t}
       \sum_{n=0}^\infty
       \frac1{n!}\left(\frac{\lambda^2t}{\lambda-\omega}\right)^n\\
   &=M\exp\left(\frac{\lambda\omega}{\lambda-\omega}t\right).
\end{aligned}
\]
Here $M\ge1$ may be assumed, so the $n=0$ term is covered by the same bound.
For every fixed $t_0>0$, the right-hand side is bounded uniformly for
$0\le t\le t_0$ and all sufficiently large $\lambda$.

We now show convergence of the generators on the original domain.  If
$y\in\Dom(A)$, then
\[
  \lambda R_\lambda y-y=R_\lambda Ay,
  \qquad
  \|R_\lambda Ay\|
  \le\frac{M}{\lambda-\omega}\|Ay\|\longrightarrow0.
\]
Since $\Dom(A)$ is dense and
$\|\lambda R_\lambda\|\le M\lambda/(\lambda-\omega)$ is uniformly bounded
for large $\lambda$, it follows that
$\lambda R_\lambda y\to y$ for every $y\in X$.  Hence, for
$x\in\Dom(A)$,
\[
  B_\lambda x=\lambda R_\lambda Ax\longrightarrow Ax.
\]
The resolvent identity makes all $B_\lambda$ commute.  Thus, for
$x\in\Dom(A)$,
\[
  T_\lambda(t)x-T_\mu(t)x
  =\int_0^t
     T_\lambda(s)T_\mu(t-s)(B_\lambda-B_\mu)x\,\dd s.
\]
The uniform semigroup bounds on compact time intervals and
$B_\lambda x\to Ax$ show that $(T_\lambda(t)x)$ is Cauchy uniformly for
$t$ in compact intervals.  Density of $\Dom(A)$ and the same uniform bounds
extend the conclusion to every $x\in X$.  Define
\[
  T(t)x=\lim_{\lambda\to\infty}T_\lambda(t)x.
\]
As in the contraction proof, uniform convergence on compact time intervals
allows passage to the limit in the semigroup law.  It also gives strong
continuity: first for $x\in\Dom(A)$ by integrating $T_\lambda(s)B_\lambda x$,
and then for general $x$ by density.  Finally,
\[
  \|T(t)\|
  \le\liminf_{\lambda\to\infty}
      M\exp\left(\frac{\lambda\omega}{\lambda-\omega}t\right)
  =Me^{\omega t}.
\]

Let $G$ be the generator of $T$.  For $x\in\Dom(A)$, dominated convergence in
\[
  T_\lambda(t)x-x=\int_0^tT_\lambda(s)B_\lambda x\,\dd s
\]
gives
\[
  T(t)x-x=\int_0^tT(s)Ax\,\dd s.
\]
Thus $A\subseteq G$.  Choose any $\mu>\omega$.  The growth estimate implies
$\mu\in\rho(G)$ by the Laplace-transform formula, while the hypothesis gives
$\mu\in\rho(A)$.  If $y\in X$ and $x=R(\mu,A)y$, then
$x\in\Dom(A)\subseteq\Dom(G)$ and
\[
  (\mu I-G)x=(\mu I-A)x=y.
\]
Hence $R(\mu,G)=R(\mu,A)$, so $G=A$.  This completes the sufficiency proof.
\end{proof}

\section{Examples}

\begin{notetheorem}[Translation semigroup]
Let
\[
  X=C_0([0,\infty))
  =\{f\in C([0,\infty)):f(t)\to0\text{ as }t\to\infty\}
\]
with the supremum norm, and define
\[
  (T(t)f)(x)=f(x+t).
\]
Then $(T(t))$ is a contraction $C_0$-semigroup.  Its generator is
\[
  Af=f',
\]
with domain
\[
  \Dom(A)=\{f\in C_0([0,\infty)):
               f\text{ is }C^1\text{ and }f'\in C_0([0,\infty))\}.
\]
\end{notetheorem}

\begin{proof}
Uniform continuity of every $f\in C_0([0,\infty))$ gives strong continuity.
For $f\in\Dom(A)$, the difference quotient converges uniformly to $f'$.  For
the converse, suppose
\[
  \frac{T(h)f-f}{h}\longrightarrow g
  \quad\text{uniformly as }h\downarrow0.
\]
Fix $x\ge0$ and $t>0$, and divide $[0,t]$ into $n$ equal pieces of length
$h=t/n$.  Telescoping gives
\[
\begin{aligned}
  f(x+t)-f(x)
  &=\sum_{k=0}^{n-1}\bigl(f(x+(k+1)h)-f(x+kh)\bigr)\\
  &=h\sum_{k=0}^{n-1}
      \frac{f(x+kh+h)-f(x+kh)}{h}.
\end{aligned}
\]
The difference quotients converge to $g$ uniformly in their spatial variable,
so the last Riemann sum converges, as $n\to\infty$, to
\[
  f(x+t)-f(x)=\int_0^t g(x+s)\,\dd s,
\]
because $g$ is continuous as a uniform limit of continuous functions.  Hence
$f$ is right differentiable at every $x\ge0$ and $f'(x)=g(x)$; at interior
points the same identity with nearby endpoints gives the ordinary two-sided
derivative.  Thus $f\in C^1$, with $f'=g\in C_0$.
\end{proof}

\begin{noteproposition}[Positive self-adjoint operators]
Let $H$ be a Hilbert space and let $A$ be densely defined, self-adjoint, and
positive:
\[
  \inner{Ax}{x}\ge0
  \qquad(x\in\Dom(A)).
\]
Then $-A$ generates a contraction $C_0$-semigroup.
\end{noteproposition}

\begin{proof}
For $x\in\Dom(A)$ and $\lambda>0$,
\[
\begin{aligned}
  \|(\lambda I+A)x\|\,\|x\|
   &\ge\operatorname{Re}\inner{(\lambda I+A)x}{x}\\
   &=\lambda\|x\|^2+\inner{Ax}{x}
    \ge\lambda\|x\|^2.
\end{aligned}
\]
Thus $\lambda I+A$ is injective, bounded below by $\lambda$, and has closed
range.  If $y$ is orthogonal to this range, self-adjointness gives
\[
  y\in\Dom((\lambda I+A)^*)=\Dom(A)
  \quad\text{and}\quad
  0=(\lambda I+A)^*y=(\lambda I+A)y,
\]
so the same lower bound forces $y=0$.  The range is therefore dense and closed,
hence all of $H$.  It follows that
\[
  \|\lambda(\lambda I+A)^{-1}\|\le1.
\]
The contraction Hille--Yosida theorem, applied to $-A$, now yields a
contraction semigroup.  The sign is essential: a positive operator itself
generally produces growth rather than contraction.
\end{proof}

\section{Dissipative operators and Lumer--Phillips}

\begin{notedefinition}
For $x\in X$, the normalized duality set is
\[
  \mathcal J(x)
  =\{x^*\in X^*:x^*(x)=\|x\|^2,
                    \ \|x^*\|=\|x\|\}.
\]
It is nonempty by Hahn--Banach.  A densely defined operator $A$ is
\emph{accretive} if, for every $x\in\Dom(A)$, there is
$x^*\in\mathcal J(x)$ with
\[
  \operatorname{Re}x^*(Ax)\ge0.
\]
It is \emph{dissipative} if $-A$ is accretive.
\end{notedefinition}

\begin{noteproposition}
An operator $A$ is dissipative if and only if
\[
  \|(\lambda I-A)x\|\ge\lambda\|x\|
  \qquad(x\in\Dom(A),\ \lambda>0).
\]
\end{noteproposition}

\begin{proof}
Suppose first that $A$ is dissipative.  Choose $x^*\in\mathcal J(x)$ with
$\operatorname{Re}x^*(Ax)\le0$.  Then
\[
\begin{aligned}
 \|(\lambda I-A)x\|\,\|x\|
  &\ge\operatorname{Re}x^*((\lambda I-A)x)\\
  &=\lambda\|x\|^2-\operatorname{Re}x^*(Ax)
   \ge\lambda\|x\|^2,
\end{aligned}
\]
which proves the norm inequality.

Conversely, assume the inequality.  For $t>0$, take $\lambda=1/t$ to obtain
\[
  \|x-tAx\|\ge\|x\|.
\]
Assume $x\ne0$.  For each $t>0$, Hahn--Banach gives $f_t\in X^*$ such that
\[
  \|f_t\|=1,
  \qquad
  f_t(x-tAx)=\|x-tAx\|;
\]
in the complex case the functional is multiplied by a unimodular scalar so
that the displayed value is real and nonnegative.  Then
\[
\begin{aligned}
  \|x\|
  &\le\|x-tAx\|\\
  &=\operatorname{Re}f_t(x)-t\operatorname{Re}f_t(Ax)\\
  &\le\|x\|-t\operatorname{Re}f_t(Ax),
\end{aligned}
\]
and hence $\operatorname{Re}f_t(Ax)\le0$.

Let $t$ tend to zero through any sequence.  Banach--Alaoglu gives a weak-$*$
convergent subnet, still denoted $f_t$, with limit $f\in B_{X^*}$.  Moreover,
\[
  |f_t(x)-\|x\||
  \le |f_t(x)-f_t(x-tAx)|
      +\bigl|\|x-tAx\|-\|x\|\bigr|
  \le 2t\|Ax\|,
\]
so $f(x)=\|x\|$.  It follows that $\|f\|=1$, and weak-$*$ convergence also
gives
\[
  \operatorname{Re}f(Ax)\le0.
\]
Thus $x^*=\|x\|f$ belongs to $\mathcal J(x)$ and verifies dissipativity.  The
case $x=0$ is immediate.
\end{proof}

\begin{notetheorem}[Lumer--Phillips theorem]
Let $A$ be densely defined on a Banach space.  Then $A$ generates a contraction
$C_0$-semigroup if and only if
\begin{enumerate}
  \item $A$ is dissipative;
  \item $\Ran(\lambda_0I-A)=X$ for some $\lambda_0>0$.
\end{enumerate}
In this case the range condition holds for every $\lambda>0$.
\end{notetheorem}

\begin{proof}
If $A$ generates a contraction semigroup, the positive resolvent is onto by
the Laplace formula.  For $x\in\Dom(A)$ and $x^*\in\mathcal J(x)$,
\[
  \operatorname{Re}x^*(T(t)x-x)
  \le \|x\|\,\|T(t)x\|-\|x\|^2\le0.
\]
Divide by $t$ and let $t\downarrow0$ to obtain
$\operatorname{Re}x^*(Ax)\le0$, so $A$ is dissipative.

Conversely, assume dissipativity and
$\Ran(\lambda_0I-A)=X$.  The norm inequality gives injectivity of
$\lambda I-A$ for every $\lambda>0$ and
\[
  \|(\lambda I-A)^{-1}\|\le\frac1\lambda
\]
on its range.  In particular, $R(\lambda_0,A)$ is a bounded operator on all of
$X$.  This also proves that $A$ is closed: if $x_n\to x$ and $Ax_n\to y$, then
\[
  x_n=R(\lambda_0,A)(\lambda_0x_n-Ax_n)
  \longrightarrow R(\lambda_0,A)(\lambda_0x-y),
\]
so $x$ belongs to $\Dom(A)$ and $Ax=y$.

Surjectivity propagates along the positive axis.  If
$\mu I-A$ is onto, then for $|\lambda-\mu|<\mu$,
\[
  \lambda I-A
  =\bigl(I+(\lambda-\mu)R(\mu,A)\bigr)(\mu I-A),
\]
and
$\|(\lambda-\mu)R(\mu,A)\|<1$.  The Neumann series therefore makes
$\lambda I-A$ onto.  Starting from $\lambda_0$, these overlapping intervals
cover $(0,\infty)$ (for example, repeatedly move from $\mu$ to $3\mu/2$).
Thus
\[
  (0,\infty)\subseteq\rho(A),
  \qquad
  \|\lambda R(\lambda,A)\|\le1.
\]
The contraction Hille--Yosida theorem now shows that $A$ generates a
contraction semigroup.
\end{proof}

\begin{notecorollary}
Let $A$ be densely defined and closed on a Banach space.  If both $A$ and its
Banach adjoint $A^*$ are dissipative, then $A$ generates a contraction
$C_0$-semigroup.
\end{notecorollary}

\begin{proof}
Dissipativity makes $\Ran(I-A)$ closed.  If it were not dense, Hahn--Banach
would give $0\ne f\in X^*$ annihilating it.  Then $f\in\Dom(A^*)$ and
$(I-A^*)f=0$, contradicting the dissipative inequality for $A^*$.  Hence the
range is dense and closed, so it equals $X$.  Apply Lumer--Phillips.
\end{proof}

\begin{notedefinition}
Let $A$ be closed.  A subspace $D\subseteq\Dom(A)$ is a \emph{core} for $A$ if
the graph of $A|_D$ is dense in $\Graph(A)$, equivalently, if $D$ is dense in
$\Dom(A)$ for the graph norm $\|x\|+\|Ax\|$.
\end{notedefinition}

\begin{notetheorem}[Invariant cores]
Let $A$ generate a $C_0$-semigroup $(T(t))$.  Suppose
$D\subseteq\Dom(A)$ is dense in $X$ and $T(t)D\subseteq D$ for every $t\ge0$.
Then $D$ is a core for $A$.
\end{notetheorem}

\begin{proof}
Choose $\lambda$ larger than a growth bound of the semigroup, so that
\[
  R(\lambda,A)x=\int_0^\infty e^{-\lambda t}T(t)x\,\dd t.
\]
Let $\overline{A|_D}$ denote the graph closure of the restriction.  We first
show that $R(\lambda,A)d$ belongs to its domain for every $d\in D$.  Approximate
the Laplace integral by finite Riemann sums
\[
  z_n=\sum_j c_{n,j}T(t_{n,j})d.
\]
Invariance gives $z_n\in D$.  Since $d\in\Dom(A)$ and
$AT(t)d=T(t)Ad$, the same Riemann sums after applying $A$ converge to
\[
  \int_0^\infty e^{-\lambda t}T(t)Ad\,\dd t
  =AR(\lambda,A)d.
\]
Thus $z_n\to R(\lambda,A)d$ in the graph norm, as required.

Now let $x\in X$ and choose $d_n\in D$ with $d_n\to x$.  The resolvent is
bounded from $X$ into $\Dom(A)$ equipped with the graph norm, because
\[
  AR(\lambda,A)=\lambda R(\lambda,A)-I.
\]
Hence
\[
  R(\lambda,A)d_n\longrightarrow R(\lambda,A)x
  \quad\text{in the graph norm}.
\]
The graph closure of $A|_D$ therefore contains $R(\lambda,A)X=\Dom(A)$.
The reverse inclusion is automatic because $A$ is closed.  Thus
$\overline{A|_D}=A$, and $D$ is a core.
\end{proof}

\begin{notetheorem}[Gaussian heat semigroup]
On $X=C_0(\mathbb R^d)$, define for $t>0$
\[
  (T(t)f)(x)
  =\frac1{(4\pi t)^{d/2}}
     \int_{\mathbb R^d}
       e^{-|x-y|^2/(4t)}f(y)\,\dd y,
  \qquad T(0)=I.
\]
Then $(T(t))$ is a contraction $C_0$-semigroup.  Its generator is the closure
of
\[
  \Delta\colon C_c^\infty(\mathbb R^d)\to C_0(\mathbb R^d),
\]
and $C_c^\infty(\mathbb R^d)$ is a core.
\end{notetheorem}

\begin{proof}
Let
\[
  G_t(x)=(4\pi t)^{-d/2}e^{-|x|^2/(4t)}.
\]
The kernels are nonnegative, have integral one, and satisfy
$G_t*G_s=G_{t+s}$; the last identity follows by completing the square (or by
Fourier transform).  Therefore $T(t+s)=T(t)T(s)$ and
\[
  \|T(t)f\|_\infty\le\|f\|_\infty.
\]
Each $T(t)f$ belongs to $C_0$: continuity follows by translating the kernel,
and vanishing at infinity follows first for compactly supported $f$ and then
for general $f\in C_0$ by uniform approximation.  Since $(G_t)$ is an
approximate identity, $T(t)f\to f$ uniformly for $f\in C_0$.  Thus $T$ is a
contraction $C_0$-semigroup.

Let $G$ denote its generator.  For $u\in\mathcal S(\mathbb R^d)$,
differentiation under the integral (equivalently, Fourier transform) gives
\[
  \frac{\dd}{\dd t}T(t)u
  =\Delta T(t)u=T(t)\Delta u.
\]
Hence $\mathcal S\subseteq\Dom(G)$ and $Gu=\Delta u$ there.  The Schwartz
space is dense in $C_0$ and invariant under $T(t)$, so the invariant-core
theorem shows that $\mathcal S$ is a core for $G$.

Finally, $C_c^\infty$ is graph dense in $\mathcal S$ for the operator
$\Delta$.  Indeed, if $\chi_R(x)=\chi(x/R)$ with $\chi=1$ near the origin,
then $\chi_Ru\in C_c^\infty$ and
\[
  \chi_Ru\to u,
  \qquad
  \Delta(\chi_Ru)\to\Delta u
\]
uniformly, because
\[
  \Delta(\chi_Ru)-\Delta u
  =(\chi_R-1)\Delta u+2\nabla\chi_R\cdot\nabla u+(\Delta\chi_R)u
\]
and $u$ is Schwartz.  Consequently $C_c^\infty$ is also a core, and $G$ is
the closure of $\Delta$ on that domain.
\end{proof}

\begin{noteremark}
The same formula is not strongly continuous on all of $C_b(\mathbb R^d)$:
functions in $C_b$ need not be uniformly continuous.  The correct standard
spaces are $C_0(\mathbb R^d)$ or the bounded uniformly continuous functions.
\end{noteremark}

\begin{notetheorem}[Poisson semigroup]
Let
\[
  P_t(x)=c_d\frac{t}{(t^2+|x|^2)^{(d+1)/2}},
  \qquad t>0,
\]
where $c_d$ is chosen so that $\int_{\mathbb R^d}P_t=1$, and define
\[
  T(t)f=P_t*f,
  \qquad T(0)=I.
\]
Then $(T(t))_{t\ge0}$ is a contraction $C_0$-semigroup on
$C_0(\mathbb R^d)$.  If $A$ denotes its generator, then
\[
  \mathcal S(\mathbb R^d)\subseteq\Dom(A^2)
\]
and, for every $u\in\mathcal S(\mathbb R^d)$,
\[
  Au=-(-\Delta)^{1/2}u,
  \qquad
  A^2u=-\Delta u,
\]
where $(-\Delta)^{1/2}$ is the Fourier multiplier with symbol $|\xi|$.
\end{notetheorem}

\begin{proof}
The kernel $P_t$ is nonnegative and has integral one.  Hence, for
$f\in C_0(\mathbb R^d)$,
\[
  \|T(t)f\|_\infty
  \le \|P_t\|_1\|f\|_\infty
  =\|f\|_\infty.
\]
Convolution with an $L^1$ kernel maps $C_0$ into $C_0$.  Indeed, continuity
follows from
\[
  |(P_t*f)(x+h)-(P_t*f)(x)|
  \le \|P_t\|_1\|f(\cdot+h)-f\|_\infty,
\]
and functions in $C_0$ are uniformly continuous.  To see vanishing at
infinity, first take $f$ compactly supported and split the integral into a
large compact set and its complement; then pass to arbitrary $f\in C_0$ by
uniform approximation with compactly supported continuous functions.

With the usual Fourier-transform normalization,
\[
  \widehat{P_t}(\xi)=e^{-t|\xi|}.
\]
Consequently
\[
  \widehat{P_t*P_s}
  =e^{-(t+s)|\xi|}
  =\widehat{P_{t+s}}.
\]
Both sides belong to $L^1(\mathbb R^d)$, so uniqueness of the Fourier
transform gives $P_t*P_s=P_{t+s}$ and therefore
$T(t)T(s)=T(t+s)$.

The family $(P_t)$ is an approximate identity.  In fact, after the change of
variables $y=tz$,
\[
  \int_{|y|\ge\delta}P_t(y)\,\dd y
  =\int_{|z|\ge\delta/t}P_1(z)\,\dd z
  \longrightarrow0
  \qquad(t\downarrow0)
\]
for every $\delta>0$.  Given $f\in C_0$ and $\varepsilon>0$, choose
$\delta>0$ so that
$|f(x-y)-f(x)|<\varepsilon$ whenever $|y|<\delta$, uniformly in $x$.  Then
\[
\begin{aligned}
  |T(t)f(x)-f(x)|
  &\le \int_{|y|<\delta}P_t(y)|f(x-y)-f(x)|\,\dd y \\
  &\quad+
    \int_{|y|\ge\delta}P_t(y)|f(x-y)-f(x)|\,\dd y \\
  &\le \varepsilon
      +2\|f\|_\infty
        \int_{|y|\ge\delta}P_t(y)\,\dd y.
\end{aligned}
\]
The right-hand side tends uniformly to $\varepsilon$, and hence
$\|T(t)f-f\|_\infty\to0$.  Thus $(T(t))$ is a contraction
$C_0$-semigroup.

Now let $u\in\mathcal S(\mathbb R^d)$.  Fourier inversion gives
\[
  \frac{T(t)u-u}{t}
  =\mathcal F^{-1}\!\left(
      \frac{e^{-t|\xi|}-1}{t}\,\widehat u(\xi)
    \right).
\]
For every $t>0$,
\[
  \left|\frac{e^{-t|\xi|}-1}{t}\right|
  \le |\xi|,
\]
and $|\xi|\,|\widehat u(\xi)|$ is integrable.  Dominated convergence in
$L^1_\xi$, followed by the $L^1$ Fourier-inversion estimate, yields uniform
convergence
\[
  \frac{T(t)u-u}{t}
  \longrightarrow
  \mathcal F^{-1}(-|\xi|\widehat u)
  =-(-\Delta)^{1/2}u.
\]
Therefore $u\in\Dom(A)$ and $Au=-(-\Delta)^{1/2}u$.  Put
\[
  g(\xi)=-|\xi|\widehat u(\xi),
  \qquad Au=\mathcal F^{-1}g.
\]
Here $g\in L^1$, so $Au\in C_0$.  Moreover, Fubini's theorem is applicable
because $P_t\in L^1$ and $g\in L^1$; it gives
\[
  T(t)Au=\mathcal F^{-1}(e^{-t|\xi|}g).
\]
Consequently
\[
  \frac{T(t)Au-Au}{t}
  =\mathcal F^{-1}\!\left(
      \frac{e^{-t|\xi|}-1}{t}(-|\xi|\widehat u)
    \right).
\]
The multiplier is dominated by
$|\xi|^2|\widehat u(\xi)|\in L^1$.  Hence
\[
  \frac{T(t)Au-Au}{t}
  \longrightarrow
  \mathcal F^{-1}(|\xi|^2\widehat u)
  =-\Delta u
\]
uniformly.  Thus $u\in\Dom(A^2)$ and $A^2u=-\Delta u$.
\end{proof}
