\chapter{Sobolev Spaces}

\section{One-dimensional Sobolev spaces}

\begin{notedefinition}
Let $I\subseteq\mathbb R$ be an open interval and let $1\le p\le\infty$.  A
function $u\in L^p(I)$ belongs to $W^{1,p}(I)$ if there is $v\in L^p(I)$ such
that
\[
  \int_I u\varphi'\,\dd x
  =-\int_I v\varphi\,\dd x
  \qquad(\varphi\in C_c^\infty(I)).
\]
The function $v$ is unique almost everywhere and is denoted by $u'$.  The norm
is
\[
  \|u\|_{W^{1,p}(I)}=\|u\|_{L^p(I)}+\|u'\|_{L^p(I)}.
\]
We write $H^1(I)=W^{1,2}(I)$.
\end{notedefinition}

\begin{notetheorem}
The space $W^{1,p}(I)$ is Banach for every $1\le p\le\infty$, separable for
$1\le p<\infty$, and reflexive for $1<p<\infty$.
\end{notetheorem}

\begin{proof}
The map
\[
  u\longmapsto(u,u')
\]
is an isometry from $W^{1,p}(I)$ into $L^p(I)\times L^p(I)$.  Its range is
closed: if $u_n\to u$ and $u_n'\to v$ in $L^p$, then for every test function
\[
  \int u\varphi'
  =\lim_n\int u_n\varphi'
  =-\lim_n\int u_n'\varphi
  =-\int v\varphi.
\]
Completeness, separability, and reflexivity therefore follow from the
corresponding product-space properties of $L^p$.
\end{proof}

\begin{notetheorem}[One-dimensional fundamental theorem]
Let $u\in W^{1,p}(I)$.  There is an absolutely continuous representative
$\widetilde u$ such that
\[
  \widetilde u(x)-\widetilde u(y)
  =\int_y^x u'(t)\,\dd t
  \qquad(x,y\in I).
\]
In particular, every one-dimensional Sobolev function has a continuous
representative.
\end{notetheorem}

\begin{proof}
Fix $y_0\in I$ and define
\[
  v(x)=\int_{y_0}^x u'(t)\,\dd t.
\]
Then $v$ is absolutely continuous and has weak derivative $u'$.  Hence
$u-v$ has weak derivative zero.  The lemma below shows that $u-v$ is almost
everywhere constant, so $u$ agrees almost everywhere with $v+c$.
\end{proof}

\begin{notelemma}
If $f\in L^1_{\mathrm{loc}}(I)$ and
\[
  \int_I f\varphi'\,\dd x=0
  \qquad(\varphi\in C_c^\infty(I)),
\]
then $f$ is almost everywhere constant on $I$.
\end{notelemma}

\begin{proof}
Choose $\psi\in C_c^\infty(I)$ with $\int_I\psi=1$.  For any
$\varphi\in C_c^\infty(I)$, the function
$\varphi-(\int\varphi)\psi$ has integral zero and hence is the derivative of
a compactly supported smooth function.  Therefore
\[
  \int f\varphi
  =\left(\int f\psi\right)\left(\int\varphi\right),
\]
which says that the distribution defined by $f$ is the constant
$c=\int f\psi$.
\end{proof}

\begin{noteproposition}[Dual characterization of a weak derivative]
Let $u\in L^p(I)$ and $1<p\le\infty$.  Then $u\in W^{1,p}(I)$ if and only
if there is $C>0$ such that
\[
  \left|\int_Iu\varphi'\,\dd x\right|
  \le C\|\varphi\|_{L^{p'}(I)}
  \qquad(\varphi\in C_c^\infty(I)),
\]
with $p'=1$ when $p=\infty$.  The smallest such $C$ is
$\|u'\|_p$.
\end{noteproposition}

\begin{proof}
One direction is Hölder's inequality.  Conversely, the map
$\varphi\mapsto-\int u\varphi'$ is bounded on the dense subspace
$C_c^\infty(I)\subset L^{p'}(I)$.  Extend it to $L^{p'}$ and apply the
$L^p$ duality theorem to obtain $v\in L^p$ with
$-\int u\varphi'=\int v\varphi$.
\end{proof}

\begin{noteremark}
At $p=1$, the displayed estimate with the $L^\infty$ norm controls the
distributional derivative only as a finite measure and characterizes bounded
variation.  It does not by itself force that measure to have an $L^1$
density.  This is why the theorem is stated for $p>1$ (and for $p=\infty$ via
$(L^1)^*=L^\infty$).
\end{noteremark}

\begin{noteremark}
For $p=1$, the absolutely continuous representative is exactly the classical
one: a function belongs to $W^{1,1}(I)$ precisely when it is almost everywhere
equal to an absolutely continuous function whose classical derivative lies in
$L^1$.  Merely bounded variation is weaker; BV functions may have jump or
singular derivatives.
\end{noteremark}

\begin{notetheorem}[Lipschitz characterization]
A function $u\in L^\infty(I)$ belongs to $W^{1,\infty}(I)$ if and only if it
has a Lipschitz representative.  Moreover,
\[
  \operatorname{Lip}(u)=\|u'\|_\infty.
\]
\end{notetheorem}

\begin{proof}
The fundamental theorem gives
$|u(x)-u(y)|\le\|u'\|_\infty|x-y|$.  Conversely, a Lipschitz function is
absolutely continuous, differentiable almost everywhere, and its derivative
is bounded by its Lipschitz constant.  Integration by parts then identifies
that derivative as the weak derivative.
\end{proof}

\begin{notetheorem}[Difference-quotient criterion on $\mathbb R$]
Let $1<p<\infty$ and $u\in L^p(\mathbb R)$.  Then $u\in W^{1,p}(\mathbb R)$
if and only if there is $C>0$ such that
\[
  \|u(\cdot+h)-u\|_p\le C|h|
  \qquad(h\in\mathbb R).
\]
The least admissible $C$ equals $\|u'\|_p$.
\end{notetheorem}

\begin{proof}
If $u\in W^{1,p}$, then for the absolutely continuous representative,
\[
  u(x+h)-u(x)=\int_0^h u'(x+s)\,\dd s.
\]
Minkowski's integral inequality gives the estimate with $C=\|u'\|_p$.
Conversely, for $\varphi\in C_c^\infty(\mathbb R)$,
\[
  \int u(x)\frac{\varphi(x-h)-\varphi(x)}{h}\,\dd x
  =\int\frac{u(x+h)-u(x)}{h}\varphi(x)\,\dd x.
\]
The right-hand side is bounded by $C\|\varphi\|_{p'}$.  Letting $h\to0$ and
using the preceding dual characterization yields a weak derivative in $L^p$.
For $p=1$, the analogous uniform translation bound characterizes bounded
variation rather than $W^{1,1}$, so the endpoint cannot be inserted without
an additional absolute-continuity condition.
\end{proof}

\section{Extension and smooth approximation}

\begin{notetheorem}[Extension from an interval]
Let $I$ be an interval and $1\le p\le\infty$.  There is a bounded linear
operator
\[
  E\colon W^{1,p}(I)\to W^{1,p}(\mathbb R)
\]
such that $Eu=u$ almost everywhere on $I$ and
\[
  \|Eu\|_{W^{1,p}(\mathbb R)}
  \le C_I\|u\|_{W^{1,p}(I)}.
\]
For a bounded interval, $Eu$ may be chosen with support in a fixed bounded
neighbourhood of $I$.
\end{notetheorem}

\begin{proof}
The construction in the notes is reflection.  We give it explicitly.  For
$I=(0,\infty)$, use the absolutely continuous representative and set
\[
  (Eu)(x)=u(|x|),
  \qquad x\in\mathbb R.
\]
Then, almost everywhere,
\[
  (Eu)'(x)=
  \begin{cases}
    u'(x),&x>0,\\
    -u'(-x),&x<0,
  \end{cases}
\]
and hence, for $p<\infty$,
\[
  \|Eu\|_p^p=2\|u\|_p^p,
  \qquad
  \|(Eu)'\|_p^p=2\|u'\|_p^p.
\]
The corresponding supremum-norm estimates hold when $p=\infty$.  Translation
handles every half-line, while $I=\mathbb R$ needs no extension.

For a bounded interval, an affine change of variables reduces the problem to
$I=(0,1)$.  Define on $(-1,2)$
\[
  v(x)=
  \begin{cases}
    u(-x),&-1<x<0,\\
    u(x),&0\le x\le1,\\
    u(2-x),&1<x<2.
  \end{cases}
\]
The one-dimensional fundamental theorem shows that $v\in W^{1,p}(-1,2)$,
with derivative obtained by changing signs on the reflected pieces, and
\[
  \|v\|_{W^{1,p}(-1,2)}\le C\|u\|_{W^{1,p}(0,1)}.
\]
Choose $\eta\in C_c^\infty(-1,2)$ with $\eta=1$ on $[0,1]$, and set
$Eu=\eta v$ on $(-1,2)$ and $Eu=0$ elsewhere.  The product rule gives
\[
  (Eu)'=\eta'v+\eta v'
\]
and hence the required bound.  Every operation is linear, and the support is
contained in the fixed compact set $\supp\eta$.  Undoing the affine change of
variables completes the construction for an arbitrary bounded interval.
\end{proof}

\begin{notelemma}[Mollification of Sobolev functions]
Let $u\in W^{1,p}(\mathbb R)$ and let $\rho\in C_c^\infty(\mathbb R)$.  Then
\[
  \rho*u\in W^{1,p}(\mathbb R),
  \qquad
  (\rho*u)'=\rho*u'.
\]
\end{notelemma}

\begin{proof}
For a test function $\varphi$, Fubini's theorem and the definition of the weak
derivative give
\[
 \int(\rho*u)\varphi'
 =\int u\,(\widetilde\rho*\varphi')
 =-\int u'\,(\widetilde\rho*\varphi)
 =-\int(\rho*u')\varphi,
\]
where $\widetilde\rho(x)=\rho(-x)$.  Young's inequality gives the $L^p$
bounds.
\end{proof}

\begin{notetheorem}[Smooth approximation]
For $1\le p<\infty$, $C_c^\infty(\mathbb R)$ is dense in
$W^{1,p}(\mathbb R)$.  More generally, smooth functions are dense in
$W^{1,p}(I)$, and on a bounded interval they may be obtained by extending,
mollifying, and restricting.
\end{notetheorem}

\begin{proof}
We first work on $\mathbb R$.  Choose $\chi\in C_c^\infty(\mathbb R)$ with
$0\le\chi\le1$, $\chi=1$ on $[-1,1]$, and $\supp\chi\subset[-2,2]$, and put
$\chi_R(x)=\chi(x/R)$.  Then
\[
  \chi_Ru\longrightarrow u
  \quad\text{in }L^p,
  \qquad
  \chi_Ru'\longrightarrow u'
  \quad\text{in }L^p
\]
by dominated convergence.  Moreover,
\[
  (\chi_Ru)'=\chi_Ru'+\chi_R'u,
  \qquad
  \|\chi_R'u\|_p\le\frac{\|\chi'\|_\infty}{R}\|u\|_p\longrightarrow0.
\]
Thus $\chi_Ru\to u$ in $W^{1,p}$, and each $\chi_Ru$ has compact support.

Now fix a compactly supported $v\in W^{1,p}(\mathbb R)$.  For mollifiers
$\rho_\varepsilon$, the convolution lemma gives
\[
  (\rho_\varepsilon*v)'=\rho_\varepsilon*v',
\]
and the $L^p$ approximate-identity theorem gives
\[
  \rho_\varepsilon*v\to v,
  \qquad
  \rho_\varepsilon*v'\to v'
\]
in $L^p$.  Hence $\rho_\varepsilon*v\to v$ in $W^{1,p}$; for small
$\varepsilon$ the convolutions are still compactly supported.  A diagonal
choice of $R$ and $\varepsilon$ proves density of $C_c^\infty(\mathbb R)$.

For a general interval, extend $u$ by the preceding extension theorem,
approximate the extension by smooth functions on $\mathbb R$, and restrict
back to $I$.
\end{proof}

\section{One-dimensional Sobolev embeddings}

\begin{notetheorem}[Sobolev embedding on an interval]
Let $I$ be an interval of finite length and $1\le p\le\infty$.  Every
$u\in W^{1,p}(I)$ has a continuous representative and
\[
  \|u\|_{L^\infty(I)}
  \le C_I\|u\|_{W^{1,p}(I)}.
\]
Thus $W^{1,p}(I)\hookrightarrow L^\infty(I)$ continuously.

If $I$ is bounded, then
\begin{enumerate}
  \item $W^{1,p}(I)\hookrightarrow C(\overline I)$ compactly for
        $1<p\le\infty$;
  \item $W^{1,1}(I)\hookrightarrow L^2(I)$ compactly.
\end{enumerate}
\end{notetheorem}

\begin{proof}
Let $L=|I|$.  We use the absolutely continuous representative of $u$.
First suppose $1\le p<\infty$.  Since $|u|^p$ is integrable, there is a point
$y\in\overline I$ such that
\[
  |u(y)|^p\le \frac1L\|u\|_p^p.
\]
For $p=1$, the fundamental theorem immediately gives, for every
$x\in\overline I$,
\[
  |u(x)|
  \le |u(y)|+\int_I|u'(t)|\,\dd t
  \le \frac1L\|u\|_1+\|u'\|_1.
\]
Now let $1<p<\infty$.  The function $|u|^p$ is absolutely continuous and,
almost everywhere,
\[
  \left|\frac{\dd}{\dd t}|u(t)|^p\right|
  \le p|u(t)|^{p-1}|u'(t)|;
\]
in the complex case this follows from
$\frac{\dd}{\dd t}|u|^p
 =p|u|^{p-2}\operatorname{Re}(\overline u\,u')$ at points where $u\ne0$,
with the usual continuous interpretation at $u=0$.  Therefore
\[
\begin{aligned}
  |u(x)|^p
  &\le |u(y)|^p
       +p\int_I|u|^{p-1}|u'| \\
  &\le \frac1L\|u\|_p^p
       +p\|u\|_p^{p-1}\|u'\|_p.
\end{aligned}
\]
Young's inequality,
$p a^{p-1}b\le(p-1)a^p+b^p$, now yields
\[
  \|u\|_\infty^p
  \le \left(\frac1L+p-1\right)\|u\|_p^p+\|u'\|_p^p,
\]
which is the desired continuous embedding.  For $p=\infty$ the conclusion is
immediate from the definition of the $W^{1,\infty}$ norm.

Let next $I$ be bounded.  If $1<p<\infty$, Hölder's inequality and the
fundamental theorem give
\[
  |u(x)-u(y)|
  \le \int_x^y|u'(t)|\,\dd t
  \le \|u'\|_p|x-y|^{1/p'}.
\]
For $p=\infty$ the same estimate holds with exponent one.  Thus a bounded set
in $W^{1,p}(I)$ is uniformly bounded and equicontinuous on $\overline I$;
Arzelà--Ascoli gives relative compactness in $C(\overline I)$.

For $p=1$, extend each function by the bounded extension operator to one fixed
bounded interval containing $\overline I$.  A bounded family is uniformly
bounded in both $L^1$ and $L^\infty$, and the fundamental theorem gives the
uniform translation estimate
\[
  \|u(\cdot+h)-u\|_1\le |h|\,\|u'\|_1
\]
(up to a harmless fixed constant caused by the extension).  The
Fréchet--Kolmogorov criterion therefore gives relative compactness in $L^1$.
If $u_n\to u$ in $L^1$ along a subsequence and
$\sup_n\|u_n\|_\infty\le M$, pass once more to a subsequence converging
almost everywhere.  Then $|u|\le M$ almost everywhere, so
\[
  \|u_n-u\|_2^2
  \le \|u_n-u\|_\infty\|u_n-u\|_1
  \le 2M\|u_n-u\|_1\longrightarrow0.
\]
Hence the embedding into $L^2(I)$ is compact.
\end{proof}

\begin{notecorollary}[Decay on an unbounded interval]
Let $I$ be unbounded and let $1\le p<\infty$.  If $u\in W^{1,p}(I)$, then its
continuous representative satisfies
\[
  u(x)\longrightarrow0
\]
at every infinite endpoint of $I$.
\end{notecorollary}

\begin{proof}
We prove the assertion at a right-hand infinite endpoint; the left-hand case
is identical after reflection.

Suppose first that $p=1$.  For $x>y$ in $I$, the fundamental theorem gives
\[
  |u(x)-u(y)|\le\int_y^x|u'(t)|\,\dd t.
\]
Because $u'\in L^1(I)$, the right-hand side tends to zero uniformly when
$x,y\to+\infty$.  Hence $u(x)$ has a finite limit $L$ at the endpoint.  If
$L\ne0$, then $|u(x)|\ge |L|/2$ on a tail of infinite length, contradicting
$u\in L^1(I)$.  Thus $L=0$.

Now let $1<p<\infty$.  Hölder's inequality gives the global modulus of
continuity
\[
  |u(x)-u(y)|
  \le \|u'\|_{L^p(I)}|x-y|^{1/p'}.
\]
Thus $u$ is uniformly continuous.  If $u(x)$ did not tend to zero, there
would be an $\varepsilon>0$ and a sequence $x_n\to+\infty$ with
$|u(x_n)|\ge\varepsilon$.  Uniform continuity supplies $\delta>0$ such that
$|u(x)-u(x_n)|<\varepsilon/2$ whenever $|x-x_n|<\delta$.  Passing to a
subsequence, the intervals
$(x_n-\delta/2,x_n+\delta/2)$ are pairwise disjoint and contained in $I$.
On each of them $|u|\ge\varepsilon/2$, so
\[
  \int_I|u|^p
  \ge \sum_n \delta\left(\frac\varepsilon2\right)^p
  =\infty,
\]
a contradiction.  Therefore $u(x)\to0$.
\end{proof}

\begin{noteremark}
On an unbounded interval, the embedding
$W^{1,p}(I)\hookrightarrow L^\infty(I)$ is not compact.  Translates of one
fixed nonzero compactly supported smooth function are bounded in $W^{1,p}$ but
have no uniformly convergent subsequence.
\end{noteremark}

\section{Product and chain rules}

\begin{noteproposition}[One-dimensional differentiation rules]
Let $1\le p\le\infty$.
\begin{enumerate}
  \item If $u,v\in W^{1,p}(I)\cap L^\infty(I)$, then
        $uv\in W^{1,p}(I)$ and
        \[
          (uv)'=u'v+uv'.
        \]
        The integration-by-parts identity
        \[
          \int_a^b u'v
          =u(b)v(b)-u(a)v(a)-\int_a^buv'
        \]
        holds for bounded $I=(a,b)$.
  \item If $u$ is real valued, $G\in C^1(\mathbb R)$,
        $G(0)=0$, and $G'$ is bounded, then
        $G\circ u\in W^{1,p}(I)$ and
        \[
          (G\circ u)'=(G'\circ u)u'.
        \]
\end{enumerate}
\end{noteproposition}

\begin{proof}
Take the absolutely continuous representatives furnished by the
one-dimensional fundamental theorem.  On every compact subinterval, the
classical product rule for absolutely continuous functions gives
\[
  (uv)'=u'v+uv'
  \quad\text{almost everywhere}.
\]
The assumption $u,v\in L^\infty$ makes both terms on the right belong to
$L^p(I)$, while $uv\in L^p(I)$.  Thus the local identity is a global weak
identity, and $uv\in W^{1,p}(I)$.  If $I=(a,b)$ is bounded, the product $uv$
is absolutely continuous on $[a,b]$, so the ordinary fundamental theorem gives
\[
  \int_a^b(u'v+uv')\,\dd x=u(b)v(b)-u(a)v(a),
\]
which is the stated integration-by-parts formula.

For the chain rule, assume $u$ is real valued, as in the statement.  Since
$G'$ is bounded and $G(0)=0$, the mean-value theorem gives
$|G(s)|\le\|G'\|_\infty|s|$, so $G\circ u\in L^p(I)$.  The composition of an
absolutely continuous function with a $C^1$ Lipschitz function is absolutely
continuous, and the classical chain rule holds almost everywhere:
\[
  (G\circ u)'=(G'\circ u)u'.
\]
The right-hand side belongs to $L^p$, proving the Sobolev chain rule.
\end{proof}

\section{Higher-order spaces, traces, and negative norms}

\begin{notedefinition}
For an integer $m\ge1$,
\[
  W^{m,p}(I)
  =\{u\in L^p(I):u^{(j)}\in L^p(I),\ 1\le j\le m\},
\]
with norm
\[
  \|u\|_{W^{m,p}(I)}
  =\sum_{j=0}^m\|u^{(j)}\|_p.
\]
Equivalent $\ell^p$ combinations of these terms give the same topology.
When $p=2$ we write $H^m(I)=W^{m,2}(I)$.
\end{notedefinition}

\begin{notedefinition}
For $1\le p<\infty$,
\[
  W_0^{1,p}(I)=\overline{C_c^\infty(I)}^{\,W^{1,p}(I)}.
\]
\end{notedefinition}

\begin{notetheorem}[Trace characterization]
Let $I=(a,b)$ be bounded and let $1\le p<\infty$.  Then
\[
  W_0^{1,p}(I)
  =\{u\in W^{1,p}(I):u(a)=u(b)=0\},
\]
where the boundary values are taken from the continuous representative.
\end{notetheorem}

\begin{proof}
The Sobolev estimate makes the trace map
\[
  u\longmapsto(u(a),u(b))
\]
continuous on $W^{1,p}(I)$.  Every function in $C_c^\infty(I)$ has zero trace,
so the closure of that space is contained in the right-hand side.

Conversely, let $u\in W^{1,p}(I)$ with $u(a)=u(b)=0$.  Choose smooth cutoffs
$\eta_\varepsilon$ such that
\[
  0\le\eta_\varepsilon\le1,
  \qquad
  \eta_\varepsilon=0
  \text{ when }\dist(x,\partial I)\le\varepsilon,
  \qquad
  \eta_\varepsilon=1
  \text{ when }\dist(x,\partial I)\ge2\varepsilon,
\]
and $|\eta_\varepsilon'|\le C/\varepsilon$.  Put
$u_\varepsilon=\eta_\varepsilon u$.  The $L^p$ tails of $u$ and $u'$ near the
endpoints tend to zero, so
\[
  (1-\eta_\varepsilon)u\to0,
  \qquad
  (1-\eta_\varepsilon)u'\to0
  \quad\text{in }L^p.
\]
It remains to control $\eta_\varepsilon'u$.  Near $a$, the zero trace and the
fundamental theorem give
\[
  |u(x)|\le\int_a^x|u'(t)|\,\dd t.
\]
If $p>1$, Hölder's inequality yields
\[
  |u(x)|^p
  \le (x-a)^{p-1}\int_a^{a+2\varepsilon}|u'(t)|^p\,\dd t,
\]

after which integration over $(a,a+2\varepsilon)$ gives
\[
  \int_a^{a+2\varepsilon}
       |\eta_\varepsilon'(x)u(x)|^p\,\dd x
  \le C\int_a^{a+2\varepsilon}|u'(t)|^p\,\dd t\longrightarrow0.
\]
For $p=1$, Fubini gives the endpoint estimate explicitly:
\[
\begin{aligned}
  \int_a^{a+2\varepsilon}|\eta_\varepsilon'(x)u(x)|\,\dd x
  &\le \frac C\varepsilon
       \int_a^{a+2\varepsilon}\int_a^x|u'(t)|\,\dd t\,\dd x \\
  &=\frac C\varepsilon
       \int_a^{a+2\varepsilon}(a+2\varepsilon-t)|u'(t)|\,\dd t \\
  &\le 2C\int_a^{a+2\varepsilon}|u'(t)|\,\dd t
  \longrightarrow0.
\end{aligned}
\]
The endpoint $b$ is identical.  Hence $u_\varepsilon\to u$ in $W^{1,p}$.
Each $u_\varepsilon$ has compact support in $I$; mollifying at a scale smaller
than its distance from the boundary gives elements of $C_c^\infty(I)$
converging to it in $W^{1,p}$.  A diagonal sequence proves the reverse
inclusion.
\end{proof}

\begin{notetheorem}[Poincaré inequality]
For a bounded interval $I$ and $1\le p<\infty$, there is $C_I>0$ such that
\[
  \|u\|_p\le C_I\|u'\|_p
  \qquad(u\in W_0^{1,p}(I)).
\]
The same estimate holds for zero-trace functions in $W^{1,\infty}(I)$.
Consequently, $\|u'\|_p$ is an equivalent norm on $W_0^{1,p}(I)$.
\end{notetheorem}

\begin{proof}
For the continuous representative and $u(a)=0$,
\[
  |u(x)|\le\int_a^x|u'(t)|\,\dd t.
\]
Hölder's inequality and integration in $x$ give the result.
\end{proof}

\begin{notedefinition}
For $m\ge1$, $W_0^{m,p}(I)$ is the closure of $C_c^\infty(I)$ in
$W^{m,p}(I)$.  For $1<p<\infty$, the dual of $W_0^{1,p}(I)$ is denoted
\[
  W^{-1,p'}(I),
\]
and the dual of $H_0^1(I)$ is denoted $H^{-1}(I)$.
\end{notedefinition}

\begin{notetheorem}[Representation of $W^{-1,p'}$]
Let $1<p<\infty$.  A functional $F$ belongs to $W^{-1,p'}(I)$ if and only if
there are $f_0,f_1\in L^{p'}(I)$ such that
\[
  \dual{F}{u}
  =\int_I f_0u\,\dd x+\int_I f_1u'\,\dd x
  \qquad(u\in W_0^{1,p}(I)).
\]
Moreover,
\[
  \|F\|_{W^{-1,p'}}
  =\inf\bigl\{
       \|f_0\|_{p'}+\|f_1\|_{p'}:
       (f_0,f_1)\text{ represents }F
     \bigr\}
\]
up to the harmless choice of an equivalent product norm.  If $I$ is bounded,
Poincaré's inequality allows a representation with $f_0=0$.
\end{notetheorem}

\begin{proof}
Embed $W_0^{1,p}(I)$ isometrically as the closed subspace
\[
  G=\{(u,u'):u\in W_0^{1,p}(I)\}
  \subset L^p(I)\times L^p(I).
\]
The functional $F$ induces a bounded functional on $G$.  Hahn--Banach extends
it to the product, and $L^p$ duality represents that extension by
$(f_0,f_1)\in L^{p'}\times L^{p'}$.  Conversely, Hölder's inequality makes
every such formula continuous.  On a bounded interval, use the derivative
embedding $u\mapsto u'$ and Poincaré's inequality instead.
\end{proof}

\section{Sobolev spaces in several variables}

\begin{notedefinition}
Let $\Omega\subseteq\mathbb R^N$ be open.  A function $u\in L^p(\Omega)$
belongs to $W^{1,p}(\Omega)$ if, for each $1\le i\le N$, there is
$g_i\in L^p(\Omega)$ satisfying
\[
  \int_\Omega u\,\partial_i\varphi\,\dd x
  =-\int_\Omega g_i\varphi\,\dd x
  \qquad(\varphi\in C_c^\infty(\Omega)).
\]
We write $\partial_i u=g_i$ and
$\nabla u=(\partial_1u,\ldots,\partial_Nu)$.  The norm
\[
  \|u\|_{W^{1,p}(\Omega)}
  =\|u\|_p+\sum_{i=1}^N\|\partial_i u\|_p
\]
may be replaced by any equivalent finite-dimensional combination of these
terms.
\end{notedefinition}

\begin{noteproposition}
The space $W^{1,p}(\Omega)$ is Banach for every $1\le p\le\infty$, separable
for $1\le p<\infty$, and reflexive for $1<p<\infty$.  The space
$H^1(\Omega)$ is Hilbert under
\[
  \inner{u}{v}_{H^1}
  =\int_\Omega u\overline v\,\dd x
   +\sum_{i=1}^N\int_\Omega
      \partial_i u\,\overline{\partial_i v}\,\dd x.
\]
\end{noteproposition}

\begin{proof}
Equip
\[
  E_p=L^p(\Omega)^{N+1}
\]
with the product norm corresponding to the displayed Sobolev norm, and define
\[
  J\colon W^{1,p}(\Omega)\longrightarrow E_p,
  \qquad
  Ju=(u,\partial_1u,\ldots,\partial_Nu).
\]
The map $J$ is an isometry onto its range.  We show that this range is closed.
Suppose
\[
  Ju_n\longrightarrow(u,g_1,\ldots,g_N)
  \quad\text{in }E_p.
\]
For every $\varphi\in C_c^\infty(\Omega)$ and every $i$,
\[
\begin{aligned}
  \int_\Omega u\,\partial_i\varphi\,\dd x
  &=\lim_{n\to\infty}
      \int_\Omega u_n\,\partial_i\varphi\,\dd x \\
  &=-\lim_{n\to\infty}
      \int_\Omega(\partial_i u_n)\varphi\,\dd x
   =-\int_\Omega g_i\varphi\,\dd x.
\end{aligned}
\]
Thus $u\in W^{1,p}(\Omega)$ and $\partial_i u=g_i$, so the limit lies in
$J(W^{1,p}(\Omega))$.  Since $E_p$ is Banach, its closed subspace
$J(W^{1,p}(\Omega))$ is Banach, and therefore so is $W^{1,p}(\Omega)$.

For $1\le p<\infty$, the space $L^p(\Omega)$ is separable; hence the finite
product $E_p$ and its closed subspace $J(W^{1,p})$ are separable.  For
$1<p<\infty$, $L^p(\Omega)$ is reflexive, so the finite product $E_p$ and its
closed subspace are reflexive.  Finally, for $p=2$ the displayed inner product
induces a norm equivalent to the chosen $W^{1,2}$ norm.  Completeness just
proved therefore makes $H^1(\Omega)$ a Hilbert space.
\end{proof}

\begin{noteproposition}[Closedness of weak differentiation]
Suppose $u_n\to u$ in $L^p(\Omega)$ and
$\partial_i u_n\to g_i$ in $L^p(\Omega)$ for every $i$.  Then
$u\in W^{1,p}(\Omega)$ and $\partial_i u=g_i$.  The same conclusion holds
when all convergences are weak in the corresponding $L^p$ spaces.
\end{noteproposition}

\begin{proof}
For $\varphi\in C_c^\infty(\Omega)$, strong convergence gives
\[
  \int_\Omega u\,\partial_i\varphi
  =\lim_n\int_\Omega u_n\,\partial_i\varphi
  =-\lim_n\int_\Omega(\partial_i u_n)\varphi
  =-\int_\Omega g_i\varphi.
\]
The same computation is valid under weak convergence, because
$\partial_i\varphi$ and $\varphi$ belong to the appropriate dual $L^{p'}$
space (with the usual endpoint interpretation).  This is exactly the
definition of the weak derivative.
\end{proof}

\begin{notedefinition}
For open sets $\omega,\Omega\subseteq\mathbb R^N$, the notation
$\omega\Subset\Omega$ means that $\overline\omega$ is compact and contained
in $\Omega$.
\end{notedefinition}

\begin{notelemma}[Local mollification]
If $u\in W^{1,p}(\Omega)$ and $\omega\Subset\Omega$, then for sufficiently
small $\varepsilon$,
\[
  u_\varepsilon=\rho_\varepsilon*u
\]
is smooth on $\omega$ and
\[
  \partial_i u_\varepsilon
  =\rho_\varepsilon*\partial_i u
  \quad\text{on }\omega.
\]
\end{notelemma}

\begin{proof}
Choose $\varepsilon<\dist(\omega,\partial\Omega)$.  For $x\in\omega$, the
integral defining $u_\varepsilon(x)$ uses only values of $u$ inside $\Omega$.
Differentiation of convolution makes $u_\varepsilon$ smooth.  To identify the
derivative, take $\varphi\in C_c^\infty(\omega)$ and use Fubini:
\[
\begin{aligned}
  \int_\omega u_\varepsilon\,\partial_i\varphi
   &=\int_\Omega u(y)
      \left(\int_\omega\rho_\varepsilon(x-y)
                    \partial_i\varphi(x)\,\dd x\right)\dd y\\
   &=-\int_\Omega \partial_i u(y)
      \left(\int_\omega\rho_\varepsilon(x-y)
                    \varphi(x)\,\dd x\right)\dd y\\
   &=-\int_\omega(\rho_\varepsilon*\partial_i u)(x)
                    \varphi(x)\,\dd x.
\end{aligned}
\]
Thus the two smooth functions have the same distributional derivative and are
equal pointwise.
\end{proof}

\begin{notetheorem}[Meyers--Serrin density theorem]
Let $1\le p<\infty$.  Then
\[
  C^\infty(\Omega)\cap W^{1,p}(\Omega)
\]
is dense in $W^{1,p}(\Omega)$.  If $\Omega=\mathbb R^N$, the approximants may
be chosen in $C_c^\infty(\mathbb R^N)$.  On a general open set,
$C_c^\infty(\Omega)$ is dense only in the smaller space
$W_0^{1,p}(\Omega)$, not in all of $W^{1,p}(\Omega)$.
\end{notetheorem}

\begin{proof}
Fix $u\in W^{1,p}(\Omega)$ and $\varepsilon>0$.  Choose an exhaustion by open
sets
\[
  \Omega_1\Subset\Omega_2\Subset\cdots\Subset\Omega,
  \qquad
  \bigcup_{j=1}^\infty\Omega_j=\Omega.
\]
Set
\[
  U_1=\Omega_2,
  \qquad
  U_j=\Omega_{j+1}\setminus\overline{\Omega_{j-1}}
  \quad(j\ge2).
\]
This is a locally finite open cover of $\Omega$.  Let $(\psi_j)$ be a smooth
partition of unity subordinate to it, chosen so that
$\supp\psi_j\Subset U_j$.  By the product rule,
\[
  u_j:=\psi_j u\in W^{1,p}(\Omega)
\]
and $u_j$ has compact support in $\Omega$.  Moreover,
$u=\sum_j u_j$, locally finitely, together with the corresponding identity for
weak derivatives.

Extend each $u_j$ by zero to $\mathbb R^N$.  Choose
$\delta_j>0$ smaller than the distance from $\supp u_j$ to
$\partial\Omega$ and so small that
\[
  \|\rho_{\delta_j}*u_j-u_j\|_{W^{1,p}(\mathbb R^N)}
  <\varepsilon 2^{-j}.
\]
The scales may also be chosen so that the enlarged supports remain a locally
finite family.  Define
\[
  v=\sum_{j=1}^\infty\rho_{\delta_j}*u_j.
\]
The sum is locally finite, so $v\in C^\infty(\Omega)$.  Since the series of
errors is absolutely summable in $W^{1,p}$,
\[
  \|v-u\|_{W^{1,p}(\Omega)}
  \le\sum_{j=1}^\infty
      \|\rho_{\delta_j}*u_j-u_j\|_{W^{1,p}(\mathbb R^N)}
  <\varepsilon.
\]
This proves density on an arbitrary open set.

When $\Omega=\mathbb R^N$, choose cutoffs
$\chi_R(x)=\chi(x/R)$.  Then
\[
  \chi_Ru\to u\quad\text{in }W^{1,p}(\mathbb R^N),
\]
because the tails of $u$ and $\nabla u$ tend to zero and
$\|u\nabla\chi_R\|_p\le C R^{-1}\|u\|_p$.  Mollifying the compactly supported
functions $\chi_Ru$ gives approximants in $C_c^\infty(\mathbb R^N)$.
\end{proof}

\begin{notetheorem}[Distribution and translation criteria]
Let $1<p<\infty$ and $u\in L^p(\mathbb R^N)$.  The following are equivalent:
\begin{enumerate}
  \item $u\in W^{1,p}(\mathbb R^N)$;
  \item for each $i$ there is $C_i$ such that
        \[
          \left|\int_{\mathbb R^N}u\,\partial_i\varphi\,\dd x\right|
          \le C_i\|\varphi\|_{p'}
          \qquad(\varphi\in C_c^\infty);
        \]
  \item there is $C>0$ such that
        \[
          \|u(\cdot+h)-u\|_p\le C|h|
          \qquad(h\in\mathbb R^N).
        \]
\end{enumerate}
For $p=1$, the analogous bounded distributional-derivative or translation
condition characterizes $BV$, not necessarily $W^{1,1}$.
\end{notetheorem}

\begin{proof}
If $u\in W^{1,p}$, Hölder's inequality gives (ii) with
$C_i=\|\partial_i u\|_p$.  Conversely, under (ii) the functional
\[
  \varphi\longmapsto-\int u\,\partial_i\varphi
\]
is bounded on the dense subspace $C_c^\infty\subset L^{p'}$.  Hahn--Banach and
$L^p$ duality give $g_i\in L^p$ representing it, so
$g_i=\partial_i u$ weakly.  Thus (i) and (ii) are equivalent.

For smooth $u$, the fundamental theorem along the segment from $x$ to $x+h$
gives
\[
  u(x+h)-u(x)=\int_0^1 h\cdot\nabla u(x+th)\,\dd t.
\]
Minkowski's inequality and translation invariance yield
\[
  \|u(\cdot+h)-u\|_p
  \le |h|\sum_{i=1}^N\|\partial_i u\|_p.
\]
Meyers--Serrin approximation passes this estimate to every
$u\in W^{1,p}$, proving (i)$\Rightarrow$(iii).

Finally, assume (iii), fix a coordinate vector $e_i$, and let
$\varphi\in C_c^\infty$.  A change of variables gives
\[
  \int u(x)\frac{\varphi(x-te_i)-\varphi(x)}{t}\,\dd x
  =\int\frac{u(x+te_i)-u(x)}{t}\varphi(x)\,\dd x.
\]
The right-hand side has absolute value at most $C\|\varphi\|_{p'}$.  Letting
$t\to0$ gives
\[
  \left|\int u\,\partial_i\varphi\right|
  \le C\|\varphi\|_{p'},
\]
which is (ii).
\end{proof}

\begin{noteproposition}[Differentiation rules in $\mathbb R^N$]
Let $1\le p\le\infty$.
\begin{enumerate}
  \item If $u,v\in W^{1,p}(\Omega)\cap L^\infty(\Omega)$, then
        $uv\in W^{1,p}(\Omega)$ and
        \[
          \partial_i(uv)=u\,\partial_i v+v\,\partial_i u.
        \]
  \item If $u$ is real valued, $G\in C^1(\mathbb R)$,
        $G(0)=0$, and $G'$ is bounded, then
        $G\circ u\in W^{1,p}(\Omega)$ and
        \[
          \partial_i(G\circ u)=(G'\circ u)\partial_i u.
        \]
  \item Let $\Phi\colon\Omega'\to\Omega$ be a $C^1$ bi-Lipschitz
        diffeomorphism whose derivative and inverse derivative are bounded.
        If $u\in W^{1,p}(\Omega)$, then $u\circ\Phi\in W^{1,p}(\Omega')$ and
        \[
          \nabla(u\circ\Phi)
          =D\Phi^{\mathsf T}(\nabla u)\circ\Phi
        \]
        almost everywhere.
  \item If $\eta\in C_c^\infty(\Omega)$ and
        $u\in W^{1,p}_{\mathrm{loc}}(\Omega)$, then
        $\eta u\in W^{1,p}(\Omega)$ and the product rule holds.
\end{enumerate}
\end{noteproposition}

\begin{proof}
For $p=\infty$, Sobolev functions have locally Lipschitz representatives; the
product, chain, and coordinate-change formulas then follow from the classical
almost-everywhere rules for Lipschitz functions.  We therefore give the
approximation argument for $1\le p<\infty$.

All statements are local, so fix $\omega\Subset\Omega$ and mollify inside a
slightly larger open set.  For the product rule, choose smooth
$u_n,v_n$ with
\[
  u_n\to u,
  \quad v_n\to v
  \quad\text{in }W^{1,p}(\omega)
\]
and almost everywhere, while
$\|u_n\|_\infty\le\|u\|_\infty$ and
$\|v_n\|_\infty\le\|v\|_\infty$.  The classical identity
\[
  \partial_i(u_nv_n)=u_n\partial_i v_n+v_n\partial_i u_n
\]
passes to $L^p$.  For example,
\[
\begin{aligned}
 \|u_n\partial_i v_n-u\partial_i v\|_p
 &\le\|u_n\|_\infty
       \|\partial_i v_n-\partial_i v\|_p
     +\|(u_n-u)\partial_i v\|_p,
\end{aligned}
\]
and the second term tends to zero by dominated convergence.  This proves (i)
on $\omega$, hence on $\Omega$.

For (ii), $G$ is Lipschitz because $G'$ is bounded, and $G(0)=0$ gives
$G(u)\in L^p$.  For smooth approximants $u_n$,
\[
  \partial_iG(u_n)=G'(u_n)\partial_i u_n.
\]
Now $G(u_n)\to G(u)$ in $L^p$, while
\[
\begin{aligned}
 \|G'(u_n)\partial_i u_n-G'(u)\partial_i u\|_p
 &\le\|G'\|_\infty
       \|\partial_i u_n-\partial_i u\|_p\\
 &\quad+
   \|(G'(u_n)-G'(u))\partial_i u\|_p,
\end{aligned}
\]
and the last term tends to zero by dominated convergence.  Closedness of weak
differentiation gives the chain rule.

For (iii), first take smooth $u$.  The classical chain rule gives the stated
formula, and the change-of-variables theorem plus boundedness of
$D\Phi$ and $D\Phi^{-1}$ gives
\[
  \|u\circ\Phi\|_{W^{1,p}(\Omega')}
  \le C_\Phi\|u\|_{W^{1,p}(\Omega)}.
\]
Approximate a general $u$ by smooth Sobolev functions and pass to the limit in
this estimate.  Assertion (iv) follows from (i) on a neighbourhood of
$\supp\eta$; compact support then makes all terms globally $L^p$.
\end{proof}

\begin{notedefinition}
For an integer $m\ge1$, a multi-index $\alpha$, and $1\le p\le\infty$,
\[
  W^{m,p}(\Omega)
  =\{u\in L^p(\Omega):D^\alpha u\in L^p(\Omega)
                      \text{ for all }|\alpha|\le m\}.
\]
It is equipped with
\[
  \|u\|_{W^{m,p}}
  =\sum_{|\alpha|\le m}\|D^\alpha u\|_p.
\]
Equivalent finite-dimensional combinations of these terms define the same
norm topology; for $p=2$ this is the Hilbert space
$H^m(\Omega)=W^{m,2}(\Omega)$.
\end{notedefinition}
% Source PDF p. 90 is already marked immediately above; omitting a second
% zero-height marker here prevents an otherwise blank continuation page.
